---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # fig.path = "README-"
  fig.path = "tools/README-"
)
set.seed(0)
```

# GauPro

<!-- badges: start -->
[![Travis-CI Build Status](https://travis-ci.org/CollinErickson/GauPro.svg?branch=master)](https://travis-ci.org/CollinErickson/GauPro)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/GauPro)](https://cran.r-project.org/package=GauPro)
[![codecov](https://codecov.io/gh/CollinErickson/GauPro/branch/master/graph/badge.svg)](https://app.codecov.io/gh/CollinErickson/GauPro)
[![R-CMD-check](https://github.com/CollinErickson/GauPro/workflows/R-CMD-check/badge.svg)](https://github.com/CollinErickson/GauPro/actions)
[![R-CMD-check](https://github.com/CollinErickson/GauPro/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/CollinErickson/GauPro/actions/workflows/R-CMD-check.yaml)
[![CRAN RStudio mirror downloads](https://cranlogs.r-pkg.org/badges/last-month/GauPro?color=blue)](https://r-pkg.org/pkg/GauPro)
<!-- badges: end -->

## Overview

This package allows you to fit a Gaussian process regression model to a dataset. 
A Gaussian process (GP) is a commonly used model in computer simulation.
It assumes that the distribution of any set of points is multivariate
normal.
A major benefit of GP models is that they provide uncertainty estimates along
with their predictions.


## Installation

You can install like any other package through CRAN.

```
install.packages('GauPro')
```

The most up-to-date version can be downloaded from 
my Github account.

```
# install.packages("devtools")
devtools::install_github("CollinErickson/GauPro")
```


## Example in 1-Dimension

This simple shows how to fit the Gaussian process regression model to data.

```{r, libraryGauPro}
library(GauPro)
```


```{r, fitsine}
n <- 12
x <- seq(0, 1, length.out = n)
y <- sin(6*x^.8) + rnorm(n,0,1e-1)
gp <- gpkm(x, y)
```

Plotting the model helps us understand how accurate the model is and how much
uncertainty it has in its predictions.
The green and red lines are the 95% intervals for the mean and for samples,
respectively.

```{r, plotsine}
gp$plot1D()
```

## Factor data: fitting the `diamonds` dataset

The model fit using `gpkm` can also be used with data/formula input and
can properly handle factor data.

In this example, the `diamonds` data set is fit by specifying the formula
and passing a data frame with the appropriate columns.


```{r fit_dm}
library(ggplot2)
diamonds_subset <- diamonds[sample(1:nrow(diamonds), 60), ]
dm <- gpkm(price ~ carat + cut + color + clarity + depth,
           diamonds_subset)
```

Calling `summary` on the model gives details about the model, including
diagnostics about the model fit and the relative importance of the features.

```{r summary_dm}
summary(dm)
```

We can also plot the model to get a visual idea of how each input affects the
output.

```{r plot_dm}
plot(dm)
```








## Using kernels

A key modeling decision for Gaussian process models is the choice of kernel.
The kernel determines the covariance and the behavior of the model.
The default kernel is the Matern 5/2 kernel (`Matern52`), and is a good
choice for most cases.
The Gaussian, or squared exponential, kernel (`Gaussian`) is a common choice
but often leads to a bad fit since it assumes the process the data comes
from is infinitely differentiable.
Other common choices that are available include the `Exponential`,
Matern 3/2 (`Matern32`),
Power Exponential (`PowerExp`),
`Cubic`,
Rational Quadratic (`RatQuad`),
and Triangle (`Triangle`).

These kernels only work on numeric data.
For factor data, the kernel will default to a
Latent Factor Kernel (`LatentFactorKernel`) for character and unordered factors,
or an Ordered Factor Kernel (`OrderedFactorKernel`) for ordered factors.
As long as the input is given in as a data frame and the columns have the
proper types, then the default kernel will properly handle it by applying
the numeric kernel to the numeric inputs and
the factor kernel to the factor and character inputs.

