---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # fig.path = "README-"
  fig.path = "tools/README-"
)
set.seed(0)
```

# GauPro

[![Travis-CI Build Status](https://travis-ci.org/CollinErickson/GauPro.svg?branch=master)](https://travis-ci.org/CollinErickson/GauPro)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/GauPro)](https://cran.r-project.org/package=GauPro)
[![codecov](https://codecov.io/gh/CollinErickson/GauPro/branch/master/graph/badge.svg)](https://codecov.io/gh/CollinErickson/GauPro)

## Overview

This package allows you to fit a Gaussian process to a dataset. 
A Gaussian process is a commonly used model in computer simulation.
It assumes that the distribution of any set of points is multivariate
normal with a constant mean and a correlation function.

The newest release allows you to use different kernel and
trend functions, instead of just a squared exponential 
covariance.

You should probably use a different package for your modeling,
such as laGP, mlegp, or GPfit if you are using R, or GPy if you
are using Python.

## Installation

You can install like any other package

```
install.packages('GauPro')
```

The most up-to-date version can be downloaded from 
my Github account.

```
# install.packages("devtools")
devtools::install_github("CollinErickson/GauPro")
```



## Examples in 1-Dimension

Fit a sine curve with noise.

```{r}
n <- 12
x <- matrix(seq(0,1,length.out = n), ncol=1)
y <- sin(2*pi*x) + rnorm(n,0,1e-1)
gp <- GauPro::GauPro(X=x, Z=y)
curve(gp$pred(x));points(x,y)
curve(gp$pred(x)+2*gp$pred(x,T)$se,col=2,add=T);curve(gp$pred(x)-2*gp$pred(x,T)$se,col=2,add=T)
```

This is the likelihood as a function of the log of theta. It is not convex and is difficult to optimize in general.

```{r}
curve(sapply(x, gp$deviance_theta_log),-10,10, n = 300) # deviance profile
```

Fit a sawtooth function with no noise.

```{r}
n <- 12
x <- matrix(seq(0,1,length.out = n), ncol=1)
y <- (2*x) %%1
gp <- GauPro::GauPro(X=x, Z=y)
curve(gp$pred(x));points(x,y)
curve(gp$pred(x)+2*gp$pred(x,T)$se,col=2,add=T);curve(gp$pred(x)-2*gp$pred(x,T)$se,col=2,add=T)
curve(sapply(x, gp$deviance_theta_log),-10,10, n = 300) # deviance profile
```

