---
title: "Surface derivatives of Gaussian process models"
author: "Collin Erickson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Surface derivatives of Gaussian process models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


This vignette covers the derivatives of the surface created by a Gaussian process model with respect to the spatial dimensions. 
The other vignette has derivatives of the deviance (likelihood) with respect to the parameters.

# Gradient of mean function

The mean function is 
$$ \hat{y}(x) = \hat{\mu} + r^T R^{-1}(Y - 1\hat{\mu}). $$
$r$ is the only part that depends on $x$, and is defined below, where $K$ is the correlation function.

$$ r = (r_1(x), \ldots, r_d(x))^T$$
$$ r_i(x) = K(z(x), z(u_i))$$
$$\frac{\partial \hat{y}(x)}{\partial x_i} = \frac{\partial r}{\partial x_i}^T R^{-1}(Y - 1\hat{\mu})$$

$$\frac{\partial r_j}{\partial x_i} = \frac{\partial K(z(x), z(u_j))}{\partial x_i} $$
This value depends on the correlation function $K$. This will give the vector $\frac{\partial r}{\partial x_i}$ which is calculates a single partial derivative, then the vector of these gives the gradient $\nabla_x \hat{y}(x)$ 




## Hessian

The second derivatives can be calculated similarly
$$\frac{\partial^2 \hat{y}(x)}{\partial x_i \partial x_k} = \frac{\partial^2 r}{\partial x_i \partial x_k}^T R^{-1}(Y - 1\hat{\mu})$$

Each element of this matrix is
$$\frac{\partial^2 r_j}{\partial x_i \partial x_k} = \frac{\partial^2 K(z(x), z(u_j))}{\partial x_i \partial x_k} $$

## Gaussian correlation derivative

The equations above work for any correlation function, but then we need to have the derivatives of the correlation function with respect to the spatial variables.

$$K(z(x), z(u_j)) = \exp(-\sum_{\ell =1}^d \theta_\ell (x_\ell - u_{j\ell})^2) $$

$$\frac{\partial K(z(x), z(u_j))}{\partial x_i}  = -2\theta_i (x_i - u_{ji}) \exp(-\sum_{\ell =1}^d \theta_\ell (x_\ell - u_{j\ell})^2)  $$
The second derivative with respect to the same dimension is

$$\frac{\partial^2 K(z(x), z(u_j))}{\partial x_i^2}  = (-2\theta_i  + 4\theta_i^2 (x_i - u_{ji})^2) \exp(-\sum_{\ell =1}^d \theta_\ell (x_\ell - u_{j\ell})^2)  $$

The cross derivative is 
$$\frac{\partial^2 K(z(x), z(u_j))}{\partial x_i \partial x_k}  = 4\theta_i \theta_k (x_i - u_{ji}) (x_k - u_{jk}) \exp(-\sum_{\ell =1}^d \theta_\ell (x_\ell - u_{j\ell})^2)  $$

## Gradient distribution

A big problem with using the gradient of the mean function of a GP is that it doesn't give an idea of its distribution/randomness. The gradient could be zero in a region where the surface is not flat just because we don't any information in that region yet.

Can we find the distribution of the gradient by taking a limit?

In $d$ dimensions we 















# default
Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
